---
title: "ESM 244 Lab 1"
author: "Carmen Galaz-García"
date: "1/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
```


### Read in SF trees

```{r}
sf_trees <- read_csv(here("data","sf_trees","sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for data wrangling and summary statistics using functions in the `dplyr` package.

Find the top 5 highest observations of trees by legal_status, do some wrangling, make a graph.

```{r}
top_5_status <- sf_trees %>% 
  count(legal_status)  %>% # groups data, counts and summarizes
  drop_na(legal_status) %>% # removes na value for specified variable
  rename(tree_count =n) %>%  # replaces name of column
  relocate(tree_count) %>% # relocates specified columns
  slice_max(tree_count, n=5) # identifies highest values for specified variable and keeps the top n
   
```

Make a graph of those top 5 observations by legal satatus

```{r}
ggplot(data = top_5_status,
       aes(x=fct_reorder(legal_status,tree_count), # convert to factor and order
           y = tree_count))+
  geom_col()+
  labs(x="legal status", y="tree count") +
  coord_flip() +
  theme_minimal()
```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees.

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>%  # detect substring
  select(legal_status, date, latitude,longitude)

ggplot(data = blackwood_acacia,aes(x=longitude, y=latitude) )+
  geom_point()
```


### tidyr::separate() and unite() functions 

Useful for combining or separating columns 

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into=c("spp_scientific", "spp_common"), sep="::")    # separate a column 

sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep="_cool_")
```






